--- 
title: 'Causal Models: Guide to `CausalQueries`'
author: "Macartan Humphreys and Alan Jacobs"
date: "`r Sys.Date()`"
bibliography:
- bib.bib
- packages.bib
description: Model based strategies for integrating qualitative and quantitative inferences.
documentclass: book
fontsize: 12pt
github-repo: rstudio/causalmodels
header-includes:
- \usepackage{amsmath}
- \usepackage{color}
- \usepackage{amssymb}
- \usepackage{amsfonts}
- \usepackage{dsfont}
link-citations: yes
site: bookdown::bookdown_site
biblio-style: apalike
---

```{r setup, include=FALSE}

 # - \usepackage{mathspec}

options(
  htmltools.dir.version = FALSE, formatR.indent = 2,
  width = 55, digits = 4, warnPartialMatchAttr = FALSE, warnPartialMatchDollar = FALSE

)
```

# Preface {-}

```{r, include=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("_packages_used.R")
```

**Map**

This guide is supplementary material for our book. [*Integrated Inferences*](https://integrated-inferences.github.io/) [@ii2023].

* The [first part](#cm)  of the guide provides a brief motivation of causal models.
* The [second part](#package) describes how the package works and how to use it. 
* The [third part](#applications)  illustrates applications of the package for defining and learning from a set of canonical causal models. 
* The short [last part](#notation) has a notation guide.


**Credits**

The approach used in `CausalQueries` is a generalization of the "biqq" models described in "Mixing Methods: A Bayesian Approach" [@humphreys2015mixing]. The conceptual extension makes use of work on probabilistic causal models described in Pearl's *Causality* [@pearl2009causality]. The approach to generating a generic stan function that can take data from arbitrary models was developed in key contributions by [Jasper Cooper](http://jasper-cooper.com/) and [Georgiy Syunyaev](http://gsyunyaev.com/). [Lily Medina](https://lilymedina.github.io/) did magical work pulling it all together and developing approaches to characterizing confounding and defining estimands. Clara Bicalho helped figure out a nice syntax for causal statements. Julio Solis made many key contributions figuring out how to simplify the specification of priors. Merlin Heidemanns figure out the `rstantools` integration and made myriad code improvements. Till Tietz revamped the package and improved every part of it.
